---
title: "Monte Carlo Method for Approximating Pi"
author: "Paul W. Egeler, MS"
date: 2020-03-14T15:00:00-05:00
categories: ["R"]
tags: ["pi", "reticulate", "monte carlo"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(reticulate)
set.seed(999)
```

In honor of Pi Day, I am going to walk through a common approach for
approximating this important transcendental number. This is a popular
programming exercise that shows how we can use random data to estimate $\pi$. I
find this interesting because it allows us to use basic principles of geometry
to approximate an irrational number.

Please note that there are better ways to compute $\pi$. However, it's my view
that this is the most intuitive way to do it. If you're interested in this
method or other methods of extimating $\pi$, check out the
[Wikipedia page](https://en.wikipedia.org/wiki/Approximations_of_%CF%80#Summing_a_circle's_area).

## How Does it Work?

Consider that the area of a circle is defined by $A_{circle}={\pi}r^2$. Then,
for a unit circle ($r = 1$),
$$
\begin{align}
A_{circle} &= {\pi}r^2 \\
&= {\pi}1^2 \\
&= \pi.
\end{align}
$$
Also, if we draw a unit square over the first quadrant of the circle, we can see
that the total area of the square is $1^2=1$ and the area of the circle within
the square is $A_{quadrant} = \frac{\pi}{4}$. The figure below illustrates this:
the area of the circle quadrant is depicted in blue while the total area of the
unit square is the sum of the areas of red and blue.

```{r area, echo = FALSE}
x <- seq(0, 1, 0.001)
y <- sqrt(1 - x**2)
plot(x = c(0,1), y = c(0,1), asp = 1, type = "n", xlab = "x", ylab = "y")
rect(0, 0, 1, 1, col = "red", border = "red")
polygon(c(0, x), c(0, y), col = "blue", border = "blue")
```

If we uniformly sample points within the unit square pictured above, the density
of those points are defined as
$$
\begin{align}
\rho &= \frac{points_{square}}{A_{square}} \\
&= \frac{points_{square}}{1^2} \\
&= points_{square}.
\end{align}
$$
Further, we can observe whether each point is within the quadrant since all points
within the quadrant satisfy $x^2 + y^2 \le 1$. Thus we approximate the area of the
quadrant since a uniform distribution of points gives
$\rho \approx \frac{points_{quadrant}}{A_{quadrant}}$, so $A_{quadrant} \approx \frac{points_{quadrant}}{\rho}$.
By substitution $A_{quadrant} \approx \frac{points_{quadrant}}{points_{square}}$. Then
$$
\begin{align}
\frac{\pi}{4} &= A_{quadrant} \\
\pi &= 4 \times A_{quadrant} \\
\pi &\approx 4 \times \frac{points_{quadrant}}{points_{square}}.
\end{align}
$$

## The Function

Thanks to R's vectorized operations, it is simple to generate many points, test
if they are within the area of the quadrant, and do the scalar multiplication and
division operations required for this analysis.

```{r}
approximate_pi <- function(n = 10000) {
  4 * sum(runif(n)**2 + runif(n)**2 <= 1) / n
}
approximate_pi()
```

## Accuracy

The number stabilizes after around 1000 sampled points. This approximation will
approach $\pi$ as the number of points sampled approaches infinity. Here are
numbers as a percent of the real $\pi$ from a sampling of 2 points to 8,388,608
points.

```{r}
pi_vec <- sapply(1:23, function(x) approximate_pi(2**x))
100 * pi_vec / pi
```

Plotted out, that looks like this:

```{r accuracy, echo = FALSE}
plot(
  pi_vec, 
  type = "b", 
  ylab = "Approximate value of pi", 
  xlab = "Number of points sampled [log2(x)]"
)
abline(h = pi, col = "red")
```

## Visualizing the Points

Now we can define a function to help us understand what's going on here.
The function below makes a bunch of random x/y coordinates and then determines
whether each ordered pair is inside the quadrant or not.

```{r}
generate_pi_data <- function(n) {
  dat <- data.frame(
    x = runif(n),
    y = runif(n)
  )
  dat$inside <- with(dat, x**2 + y**2 <= 1)
  dat
}

pi_data <- generate_pi_data(2**13)
```

Here are the first few rows of data.

```{r}
head(pi_data)
```

And here are the points plotted with the color coding from our first figure.

```{r points, echo=FALSE}
with(pi_data,
  plot(
    x, y,
    col = scales::alpha(ifelse(inside, "blue", "red"), 0.25),
    pch = 20L,
    asp = 1
  )
)
```

## Bonus Python Code

I've been wanting to try the `reticulate` package for including python code in
Rmarkdown reports. So here's some bonus material.

You can sum the area systematically by sampling points at set intervals. Here is
some python code to do that.

```{python}
def estimate_pi(n):
    i = 0

    for x in range(n):
        x = x / (n - 1)
        for y in range(n):
            y = y / (n - 1)
            if x*x + y*y <= 1:
                i += 1

    return 4*i/(n*n)

estimate_pi(1000)
```

Iterating over several sample sizes...

```{python}
for i in range(1, 12):
    print('% 8i: %f' % (2**(2*i), estimate_pi(2**i)))
```
